Freezing of a Weeks-Chandler-Anderson Fluid using Classical Density Functional Theory

### 2021-02-02 papers/thesis-kirstie/thesis.tex

Classically, fluids have been modeled by a collection of small, 
impenetrable spheres colliding with one another like billiard balls. 
In the simplest case, the spheres are treated as having only a repulsive 
potential that rises instantly to infinity when the spheres make contact. 
This model, called the hard-sphere fluid, has served as the 
primary reference system for fluids. This is due to the fact that the 
short-range repulsion between atoms is much stronger than their long-range 
attraction, and thus dominates the interactions between atoms.

In more recent times, other model fluids have been developed.
Zwanzig introduced the square-well 
fluid in 1954 which is a perturbation of the hard-sphere fluid created by adding 
a hard-edge short-range attractive potential to the repulsive potential of the 
hard-sphere fluid~\cite{ZwanzigSqrWell}. In 1964 Rowlinson studied 
spheres with softly fading short-range repulsive potentials which he modeled by making
the radius of hard spheres temperature dependent~\cite{rowlinson1964statistical}. 
This reflects the idea that soft-spheres penetrate more deeply into 
each other when they collide at higher temperatures. 
In 1967 Barker and Henderson combined the perturbation approach of Zwanzig 
with Rowlinson's idea of using temperature-dependent radii to model spheres 
with soft repulsive potentials using hard-spheres~\cite{barker1967perturbation}. 
In this model, 
%a soft-sphere fluid is modeled by using a hard-sphere fluid with 
hard spheres are given larger radii 
at low temperatures, and smaller radii at high temperatures. 
In 1971 Weeks, 
Chandler, and Anderson used temperature and density dependent hard-sphere 
radii to model spheres with a more realistic softly fading repulsive 
potential similar to that between neutral atoms.
%portion of the Leonard-Jones potential
Their is known as the Weeks-Chandler-Anderson (WCA) model fluid~\cite{andersen1971relationship} . 
Models continue to be developed to characterize fluids with particular 
properties, or to address problems like surface tension, 
interfaces, adsorption, wetting and confinement.
A review of some model fluids and problems that have been
studied more recently is given by L\"{o}wen~\cite{lowen}.

While models for particular fluids were being created, powerful methods 
to predict the properties of fluids were also being developed. 
In the late 1950s, Scaled Particle Theory (SPT)~\cite{ReissSPT} was 
introduced from which an equation of state, and an expression for the 
excess Helmholtz free energy of a homogeneous hard-sphere fluid can be 
obtained~\cite{Hansen}. 
In 1958, Percus-Yevick introduced a method to describe the thermodynamics 
of the hard-sphere fluid based on an approximate closure to the 
Ornstein-Zernike integral equation~\cite{PercusYevick}.
%Percus and Yevick took another approach to the hard-sphere fluid.
In 1961 Morita and Hiroike introduced a density functional 
approach~\cite{MoritaDFT} to finding the equilibrium Grand and Helmholtz free energies  
%, and the corresponding equilibrium number density profile 
of a classical inhomogeneous fluid. 
Density functional theory was further developed for inhomogeneous classical fluids
by Stillinger and Buff (1962)~\cite{stillingerandbuff}, Lebowitz and 
Percus (1963)~\cite{lobowitzandpercus}, and Mermin (1965)~\cite{merium} among others. A review
of the development of cDFT is found in Fundamentals of Inhomogeneous fluids~\cite{inhomogeneousfluids}.

To use classical Density Functional Theory (cDFT), an expression for the free energy 
as a functional of a spatially varying average single-particle number density $n(\vec r)$
%that varies spatially 
is constructed for a particular fluid.
%DFT Evans 1979
%Led by the consistency between results from SPT and Percus-Yevick, 
In 1989 Rosenfeld introduced Fundamental Measure Theory (FMT)~\cite{rosenfeld1989,rosenfeld1988} 
to form a functional for the hard-sphere fluid based on the idea that 
%the lowest term in the virial expansion of the excess chemical potential 
the lowest order term in the virial expansion of the excess chemical 
potential of a hard-sphere fluid can 
be expressed in terms of a weighted density using a weight function 
%using a weight 
%function 
of the form of 
a backward Heaviside step function
%\begin{equation}
   %F_{ex}=k_BT\int_{volume} n(\vec r) \theta(|\vec r - \vec r'|-(\vec R_i + \vec R_j)) d\vec r
%\end{equation}
that can itself be expressed in terms of a sum of convolutions involving 
a set of weight functions related to the geometry of a sphere. 
For two spheres with radii, 
$R_i$  and $R_j$, the convolution is written as~\cite{rosenfeld1989}
\begin{displaymath}
   \Theta(|\vec r - \vec r'|-(R_i + R_j)) = w_i^{(3)} \otimes w_j^{(0)} + w_j^{(3)} \otimes w_i^{(0)} + w_i^{(1)} \otimes w_j^{(2)} + w_j^{(1)} \otimes w_i^{(2)}\end{displaymath}
\begin{equation}   + \vec w_i^{(1)} \otimes \vec w_j^{(2)} + \vec w_j^{(1)} \otimes \vec w_i^{(2)}
\end{equation}

FMT gives the same result for the excess Helmholtz free energy of the 
hard-sphere fluid as SPT does, from which it can be 
constructed~\cite{rosenfeld1989,Hansen,santos2012phi3}. It also gives the same 
result for the direct correlation function approximation given by Percus-Yevick. 
However, it did not predict freezing due to one term in the expression that
diverged when the fluid was highly confined.  
Tarazona later solved this problem by introducing a tensor version of $\Phi_3$,
one of the energy density terms 
in the functional,~\cite{tarazonaphi3, santos2012phi3} 
which took into account the non-spherical, and thus 
non-symmetric, shape of the cavity in which the fluid is confined 
while also yielding the correct excess free energy for the fluid as its 
confinement is reduced from a three-dimensional volume (as in a bulk fluid), 
to a two-dimensional plane (as in a thin-film), to a one-dimensional 
line (as in a fluid confined to a capillary tube), to a zero-dimensional 
dot, as in an atom located at a lattice site within a crystal~\cite{rothreview}.

In 1999 Schmidt adopted FMT for spheres with soft 
interatomic potential energies
(which rise to infinity over a distance, rather than instantaneously 
like the potential energy between hard spheres)
forming Soft Fundamental Measure Theory (SFMT)~\cite{schmidt1999density}. 
%by replacing $w^{(3)}$ with a weight function that would result in  
%soft interatomic potentials (which rise to infinity over a distance, 
%rather than instantaneously).
Schmidt applied his Soft Fundamental Measure Theory to various fluids, 
%with soft, short-range repulsive and attractive potentials, 
and developed 
an error function model 
%based on a Gaussian weight function 
for spheres with steep, short-range soft repulsive 
potentials~\cite{schmidt2000fluid}.
%, similar to that exhibited by a WCA fluid. 
For some more recent developments in classical Density Functional Theory, 
see the review by Evans ~\cite{evansreview} (2016), the book 
Theory of Simple Fluids by Hansen and McDonald~\cite{Hansen} (2013), 
and the paper by Tschopp and coworkers~\cite{BHextended} (2020) in which 
they extend the Barker-Henderson model to inhomogeneous 
fluids from first principles.

Classical density functional theory has also been applied 
to phase transitions, and freezing. In 1979 Ramakrishnan and Yussouff studied 
freezing by viewing a 
crystal as a highly inhomogeneous liquid~\cite{ramakrishnanandyussouff}. 
In 2010 Oettel and fellow researchers studied hard-sphere crystals with 
fully developed FMT~\cite{Oettel2010}, and 
in 2016 FMT was used to find the fluid and crystal phases of the 
Asakura-Oosawa model fluid~\cite{asakuraoosawa}. 
Lutsko and Lam studied studied crystallization of the Leonard-Jones 
fluid~\cite{lutsko} in 2018, and Lutsko and Schoonen looked at cDFT applied 
to the solid state~\cite{lutskoschoonen2020} in 2020.
In all these studies the hard-sphere fluid has been used as the reference 
system to which other terms are added to create 
the full functional for the fluid being studied.

In this paper, classical Density Functional Theory will be used to predict 
the freezing of a Weeks-Chandler-Anderson model fluid directly by applying 
SFMT with Schmidt's proposed error function model, 
in which we have incorporated our temperature dependent parameters to 
fit the functional to the soft-sphere WCA fluid. 
%form the functional for the excess Helmholtz free energy. 
%for generating the weight functions, and Tarazona's tensor functional. 
Results will be compared to those generated from Monte-Carlo simulations.

%Newer Intro....
%The goal of this project is to generate temperature-density 
%and pressure-temperature phase diagrams for a Weeks-Chandler-Anderson 
%(WCA) model fluid using classical Density Functional Theory (cDFT). 
%To do this, the Helmholtz free energies of the fluid
%in a fluid state and in a crystalline solid state must be found.
%In general, the Helmholtz free energy, which consists of an ideal part 
%and an excess part, can be written as a functional of the number density 
%profile $n(\vec{r})$ which describes the way the average number 
%density varies spatially. 
%\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}

@@

In developing a model for a fluid, simplifications can be made regarding 
the composition, type, shape, and interatomic interaction of the molecules. 
Rather than dealing with any particular molecule or atom, the model only 
deals with generic ``particles", which may also be referred to as ``atoms''. 
The shape of the particles, or atoms will be the simplest shape possible - 
a sphere. The spheres in the simplest models do not exhibit quantum 
interference effects as the mass of the spheres, and the temperature 
are taken to be large enough that the de Broglie wavelength is smaller 
than the mean distance between the spheres. The fluid can then be treated 
as a classical fluid where the spheres interact with each other according 
to their interatomic potential.

@@

The combined attractive and repulsive effects produce the overall shape of 
the Lennard-Jones interatomic potential shown in Figure \ref{fig:LJ_potential}. 
The potential is zero when the distance r between the centers of two atoms 
is equal to $\sigma$. 
%At this point the attraction between the atoms just cancels their repulsion. 
As the distance 
between the centers of the two atoms decreases below $\sigma$, the potential 
rapidly approaches positive infinity corresponding to infinite repulsion. 
Above $\sigma$ the potential becomes negative indicating a net attraction 
between the atoms. The net attractive energy is maximized at a distance of 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the radius of 
the sphere given by $R={2^{-\frac{5}{6}}}\sigma$. At this point the depth of
the potential well, given by epsilon $\epsilon$, is a maximum, and the force
between the spheres is zero. As r increases 
further, the net attraction decreases and the potential approaches zero
asymptotically. The attractive and repulsive effects can be separated into two
potentials, shown with different colors in Figure \ref{fig:LJ_potential_2parts}, 
which when added together regenerate the full Lennard-Jones potential.

\begin{figure}
    \centering
    \includegraphics[height=7cm]{LJ_Potential_2part.pdf}
    \caption{Lennard-Jones potential broken into two parts}
    \label{fig:LJ_potential_2parts}
  \end{figure}

\begin{figure}
    \centering
    \includegraphics[height=7cm]{WCA_Potential.pdf}
    \caption{WCA potential}
    \label{fig:WCA_potential}
  \end{figure}

The repulsive portion of the Lennard-Jones potential forms the potential for a 
Weeks-Chandler-Anderson (WCA) model fluid~\cite{andersen1971relationship},
shown in Figure \ref{fig:WCA_potential}. 
The WCA potential is described by the equation 
\begin{align} \label{eq:VWCA}
    V_{WCA}=\left\{\begin{array}{rcl} {4\epsilon{\left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]}+\epsilon} & \mbox{for} & 0<r<{2R} \\ 0 & \mbox{for} & r>2R \end{array}\right.
\end{align} 
The WCA potential decreases as the distance between the 
centers of the atoms r increases, and becomes, and remains, essentially 
zero when r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the 
radius of the sphere. 
As with the Lenard-Jones potential, the repulsive WCA potential increases 
rapidly and approaches infinity as the two spheres begin to overlap. This 
is a more realistic repulsive potential than that of the hard-sphere model 
which goes instantly to infinity when the two spheres make contact (at r=2R), 
and is zero otherwise. Because the WCA spheres have no hard boundary in 
their interatomic potential, they are called ``soft spheres'' and their 
gradually fading potential makes them slightly ``squishy'', as pictured in
Figure~\ref{fig:TwoSpheres}.

\begin{figure}
    \centering
    \includegraphics[height=3.5cm]{figs/TwoSpheres.pdf} 
    \caption{(Left) The WCA potential is zero at r=2R. 
             (Right) The WCA potential increases rapidly for $r<2R$.}
    \label{fig:TwoSpheres}
\end{figure}

@@

%Freezing of simple fluids is governed by entropy, which the universe
%seeks to maximize according to the second law of Thermodynamics. When the 
%entropy of a pure substance in its liquid state is lower than the entropy 
%that it would have in its solid state at the same temperature (necessary 
%for thermal equilibrium with the environment), the same pressure (necessary for mechanical 
%equilibrium with the environment), and the same chemical potential (necessary for diffusive 
%equilibrium with the environment), a phase transition occurs, that is, the liquid 
%freezes. 
%%Along the phase boundary between pure liquid and pure solid states, both 
%%liquid and solid states can coexist in equilibrium. 
%Freezing of simple fluids can be seen when the entropy that a pure substance 
%would have in a liquid state is lower than the entropy 
%that it would have in its solid state at the same temperature (necessary 
%for thermal equilibrium), the same pressure (necessary for mechanical 
%equilibrium), and the same chemical potential (necessary for diffusive 
%equilibrium).

%A liquid will eventually freeze as heat is removed, or the liquid is compressed, and 
A liquid will freeze when the free energy it would have as a solid is lower 
than the free energy that it has as a liquid at the same density.
At the phase transition the density will change abruptly, and
the particular structure and density that result will be affected 
by the interatomic potential. When repulsive, soft sphere fluids 
(that are not too soft) freeze, they form lattice structures composed of 
Face Centered Cubic (FCC) crystal cells~\cite{Hansen}, as do hard spheres. 
This structure is what one would see by filling a box with marbles. It 
results in the highest density achievable for hard, and nearly-hard, spheres 
that are purely repulsive.

\begin{figure}
    \centering
    \includegraphics[height=7cm]{figs/P-T_Diagram.pdf}
    \caption{Conceptual Pressure-Temperature Phase Diagram}
    \label{fig:P-T_Diagram}
\end{figure}

A phase transition occurs along a phase boundary line on a phase diagram
such as can be seen in the pressure-temperature (P-T) phase diagram
shown in Figure~\ref{fig:P-T_Diagram}. 
%
%Points along the phase boundary that separates the solid and the fluid phases 
%indicate a phase transition where the solid and fluid phases coexist, 
%and show that both the solid and the fluid have the same temperature (necessary 
%for thermal equilibrium)
%and the same pressure (necessary for mechanical 
%equilibrium)at a phase transition, and the same chemical potential 
%(necessary for diffusive equilibrium). 
%
This phase diagram is 
representative of some substance 
which exhibits both attraction and repulsion. However, since 
the atoms in a WCA fluid do not exhibit attraction there will not be a 
dividing line between the gas and liquid phases, there will simply
be a fluid.
The phase diagram indicates the equilibrium state a substance will be in at
different temperatures and pressures.
At a phase transition, both the fluid and the solid have 
the same temperature (necessary 
for thermal equilibrium), the same pressure (necessary for mechanical 
equilibrium) and the same chemical potential (necessary for diffusive equilibrium).
%The temperature , pressure, and chemical potential will be the same for both
%the fluid and the solid at a fluid-solid phase transition.

The pressure can be related to the Helmholtz free energy F, which is minimized 
for a system at fixed temperature, volume, and number of particles, as the 
entropy of the system and its surroundings added together is maximized. The negative of 
the partial derivative of the Helmholtz free energy with respect to the volume V 
at a fixed temperature T and number of particles N, equals the pressure. 
\begin{equation}{P=-\frac{\partial{F}}{\partial{V}}\bigg|_{T,N}}\end{equation}
\noindent Dividing both the Helmholtz free energy and the volume by the 
number of particles, the partial derivative of the Helmholtz free energy 
with respect to volume becomes the partial derivative of the Helmholtz 
free energy per atom, $f$,  with respect to the inverse number density. 
\begin{equation}{P=-\frac{\partial{\frac{F}{N}}}{\partial{\frac{V}{N}}}\bigg|_{T,N} = -\frac{\partial{f}}{\partial{\frac{1}{n}}}\bigg|_{T,N}}\end{equation} 
Thus, the pressure corresponds to the negative of the slope of the curve 
on a plot of the Helmholtz free energy per atom versus the inverse 
density at fixed temperature.

 \begin{figure}
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig1.pdf}
    \caption{Fluid and solid densities at the fluid-solid phase 
    transition are identified using Maxwell's Double Tangent Construction}
    \label{fig:MaxwellDT}
  \end{figure}

Two such curves are shown in Figure~\ref{fig:MaxwellDT}, one for the fluid
and one for the solid at a fixed temperature. 
%Each curve consists of a collection of data points $(\frac{F}{N}, \frac{1}{n})$ 
The data for the curves 
%and the solid respectively, which 
will be obtained by using cDFT. 
%to find the Helmholtz free energies of the fluid and the solid for various 
%temperatures and number densities. 
Since the pressure of the 
fluid is the same as the pressure of the solid at the phase transition, 
the density of the fluid and the density of the solid at 
freezing can be found from identifying the points at which the 
tangents to each curve have the same slope, as shown by the black line. 
This process is called Maxwell's Double Tangent Construction.

To facilitate finding the pressure at the phase transition computationally, 
it is helpful to take different approach by considering the Gibbs free energy given by G=F+PV. 
Rearranging the terms and dividing by N gives an  expression for the 
Helmholtz free energy per atom $f$ in terms of the Gibbs free energy 
per atom  $g$ (which is the chemical potential)  
\begin{equation}\frac{F}{N} = -P\frac{V}{N}+\frac{G}{N} \end{equation}
\begin{align}f=-P\frac{1}{n}+g\end{align}
This equation has the form of a line
\begin{equation}
    y=mx+y_o
\end{equation}
and describes the black line shown in Figure~\ref{fig:MaxwellDT} where the 
pressure is given by the negative of the slope of the line, and the Gibb's 
free energy per atom, or chemical potential is given by the y-intercept. 
At the phase transition the chemical potential of the fluid is 
the same as the chemical potential of the solid, as is the pressure.
\begin{equation}
    \mu_F=\mu_S\mbox{~~~~and~~~~}P_F=P_S\mbox{~~~~at a phase transition}
\end{equation}

 \begin{figure}
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig2.pdf}
    \caption{Pressure is found at the point of intersection where $\mu_F=\mu_S$.}
    \label{fig:GibbsvsP}
  \end{figure}

\begin{figure}
    \centering
    \includegraphics[height=8cm]{figs/MaxwellDTC-Fig3.pdf}
    \caption{The fluid and solid densities at the phase transition are 
    found from a plot of pressure vs inverse density for both the fluid and the solid.}
    \label{fig:Pvsinvn}
  \end{figure}

>>>>

Classically, fluids have been modeled by a collection of small, 
impenetrable spheres colliding with one another like billiard balls. 
In the simplest case, the spheres are treated as having only a repulsive 
potential. This model, called the hard-sphere fluid, has served as the 
primary reference system for fluids. This is due to the fact that the 
repulsion between atoms is much stronger than their attraction, and so 
the repulsive potential, which is descriptive of hard spheres, dominates 
the interactions between atoms. In 1954 Zwanzig introduced the Square-well 
fluid which is a perturbation of the hard-sphere fluid created by adding 
a hard-edge attractive potential to the repulsive potential of the 
hard-sphere fluid~\cite{ZwanzigSqrWell}. In 1964 Rowlinson studied 
spheres with softly fading repulsive potentials which he modeled by making
the radius of the spheres temperature dependent~\cite{rowlinson1964statistical}. 
This reflects the idea that soft-spheres penetrate more deeply into 
each other when they collide at higher temperatures. 
In 1967 Barker and Henderson combined the perturbation approach of Zwanzig 
with Rowlinson's idea of using temperature-dependent radii to model spheres 
with soft repulsive potentials to create the Barker-Henderson model 
fluid~\cite{barker1967perturbation}. In this model, a soft-sphere fluid is 
modeled by using a hard-sphere fluid with hard spheres of larger radii 
at low temperatures, and smaller radii at high temperatures. In 1971 Weeks, 
Chandler, and Anderson used temperature and density dependent hard-sphere 
radii to model spheres with the more realistic softly fading repulsive 
portion of the Leonard-Jones potential~\cite{andersen1971relationship}, 
which is known as the Weeks-Chandler-Anderson model fluid.

ADD MORE HERE TO COME UP TO DATE!!
\newline

While models for particular fluids were being created, powerful methods 
to describe the properties of fluids were also being developed. 
In the late 1950s, Scaled Particle Theory (SPT)~\cite{ReissSPT} was 
introduced from which an equation of state, and an expression for the 
excess Helmholtz free energy of a homogeneous hard-sphere fluid can be 
obtained~\cite{Hansen}. 
%\textcolor{red}{
In 1958, Percus-Yevick introduced an alternative method 
for the hard-sphere fluid based on an approximate closure to the 
Ornstein-Zernike integral equation~\cite{PercusYevick}.
%Percus and Yevick took another approach to the hard-sphere fluid.
In 1961 Morita and Hiroike introduced a density functional 
approach~\cite{MoritaDFT} to finding the equilibrium Grand and Helmholtz free energies, 
and the corresponding equilibrium number density profile for 
inhomogeneous fluids. To use the classical Density Functional Theory (cDFT) 
approach, an expression for the free energy 
as a functional of an average number density that varies spatially 
must be constructed for a particular fluid.
%DFT Evans 1979
%Led by the consistency between results from SPT and Percus-Yevick, 
In 1989 Rosenfeld introduced Fundamental Measure Theory (FMT)~\cite{rosenfeld1989,rosenfeld1988} 
to form a functional for the hard-sphere fluid based on the idea that 
the lowest term in the virial expansion 
of the excess chemical potential for a hard-sphere fluid can 
be expressed in terms of a sum of weighted densities using a weight 
function of the form of a backward Heavyside step function (shown below) 
that can be expressed in terms of a sum of convolutions involving 
a set of weight functions related to the geometry of a sphere.
\begin{displaymath}
   \theta(|\vec r - \vec r'|-(\vec R_i + \vec R_j)) = w_i^{(3)} \otimes w_j^{(0)} + w_j^{(3)} \otimes w_i^{(0)} + w_i^{(1)} \otimes w_j^{(2)} + w_j^{(1)} \otimes w_i^{(2)}\end{displaymath}
\begin{equation}   + \vec w_i^{(1)} \otimes \vec w_j^{(2)} + \vec w_j^{(1)} \otimes \vec w_i^{(2)}
\end{equation}
FMT gives the same result for the excess Helmholtz free energy of the hard-sphere fluid as 
SPT does, from which it can be constructed~\cite{rosenfeld1989,Hansen,santos2012phi3},
and for the direct correlation function approximation given by Percus-Yevick.
Tarazona later improved on the 
method by introducing a tensor version of one of the energy density terms 
in the functional~\cite{tarazonaphi3, 
santos2012phi3}, and in 1999 Schmidt adopted FMT for spheres with soft 
interatomic potential energies
(which rise to infinity over a distance, rather than instantaneously 
like the potential energy between hard spheres)
forming Soft Fundamental Measure Theory (SFMT)\cite{schmidt1999density}.

%by replacing $w^{(3)}$ with a weight function that would result in  
%soft interatomic potentials (which rise to infinity over a distance, 
%rather than instantaneously).

Schmidt applied his Soft Fundamental Measure Theory to various fluids 
with soft, short-range repulsive and attractive potentials, and developed 
an error function model 
%based on a Gaussian weight function 
for spheres with steep, short-range soft repulsive 
potentials~\cite{schmidt2000fluid}.
%, similar to that exhibited by a WCA fluid. 
In this paper, classical Density Functional 
Theory will be used to predict the freezing of a Weeks-Chandler-Anderson 
model fluid by applying SFMT with Schmidt's proposed error function model, 
in which we have incorporated our temperature dependent parameters to fit the 
functional to the soft-sphere WCA fluid.
%form the functional for the excess Helmholtz free energy. 
%for generating the weight functions, and Tarazona's tensor functional. 
Results will be compared to those generated from Monte-Carlo simulations.

%Results will be compared to those from Monte-Carlo simulations.

%Newer Intro....
%The goal of this project is to generate temperature-density 
%and pressure-temperature phase diagrams for a Weeks-Chandler-Anderson 
%(WCA) model fluid using classical Density Functional Theory (cDFT). 
%To do this, the Helmholtz free energies of the fluid
%in a liquid state and in a crystalline solid state must be found.
%In general, the Helmholtz free energy, which consists of an ideal part 
%and an excess part, can be written as a functional of the number density 
%profile $n(\vec{r})$ which describes the way the average number 
%density varies spatially. 
%\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation}

@@

In developing a model for a fluid, simplifications can be made regarding 
the composition, type, shape, and interatomic interaction of the molecules. 
Rather than dealing with any particular molecule or atom, the model only 
deals with generic ``particles", which may also be referred to as ``atoms''. 
The shape of the particles, or atoms will be the simplest shape possible - 
a sphere. The spheres in the simplest models do not exhibit quantum 
interference effects as the mass of the spheres, and the temperature 
are taken to be large enough that the de Broglie wavelength is smaller 
than the mean distance between the spheres. The fluid can then be treated 
as a classical fluid where the spheres interact with each other according 
to their electrostatic interatomic potential.

@@

The combined attractive and repulsive effects produce the overall shape of 
the Lennard-Jones interatomic potential shown in Figure \ref{fig:LJ_potential}. 
The potential is zero when the distance r between the centers of two atoms 
is equal to $\sigma$. 
%At this point the attraction between the atoms just cancels their repulsion. 
As the distance 
between the centers of the two atoms decreases below $\sigma$, the potential 
rapidly approaches positive infinity corresponding to infinite repulsion. 
Above $\sigma$ the potential becomes negative indicating a net attraction 
between the atoms. The net attraction is maximized at a distance of 
r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the radius of 
the sphere given by $R={2^{-\frac{5}{6}}}\sigma$. At this point the depth of
the potential well, given by epsilon $\epsilon$, is a maximum. As r increases 
further, the net attraction decreases and the potential approaches zero
asymptotically. The attractive and repulsive effects can be separated into two
potentials, shown with different colors in Figure \ref{fig:LJ_potential_2parts}, 
which when added together regenerate the full Lennard-Jones potential.

\begin{figure}
    \centering
    \includegraphics[height=7cm]{LJ_Potential_2part.pdf}
    \caption{Lennard-Jones potential broken into two parts for $\sigma=1$, $\epsilon=1$}
    \label{fig:LJ_potential_2parts}
  \end{figure}

\begin{figure}
    \centering
    \includegraphics[height=7cm]{WCA_Potential.pdf}
    \caption{WCA potential for $\sigma=1$, $\epsilon=1$}
    \label{fig:WCA_potential}
  \end{figure}

The repulsive portion of the Lennard-Jones potential forms the potential for a 
Weeks-Chandler-Anderson (WCA) model fluid~\cite{andersen1971relationship},
shown in Figure \ref{fig:WCA_potential}. 
The WCA potential is described by the equation 
\begin{align} \label{eq:VWCA}
    V_{WCA}=\left\{\begin{array}{rcl} {4\epsilon{\left[\left(\frac{\sigma}{r}\right)^{12} - \left(\frac{\sigma}{r}\right)^6 \right]}+\epsilon} & \mbox{for} & 0<r<{2R} \\ 0 & \mbox{for} & r>2R \end{array}\right.
\end{align} 
The WCA potential decreases from infinity as the distance between the 
centers of the atoms r increases, and becomes, and remains, essentially 
zero when r=$2^\frac{1}{6}\sigma=2R\approx{1.122}\sigma$ where R is the 
radius of the sphere. 
As with the Lenard-Jones potential, the repulsive WCA potential increases 
rapidly and approaches infinity as the two spheres begin to overlap. This 
is a more realistic repulsive potential than that of the hard-sphere model 
which goes instantly to infinity when the two spheres make contact (at r=2R), 
and is zero otherwise. Because the WCA spheres have no hard boundary in 
their interatomic potential, they are called ``soft spheres'' and their 
gradually fading potential makes them slightly ``squishy'', as illustrated in
Figure~\ref{fig:TwoSpheres}.

\begin{figure}
    \centering
    \includegraphics[height=3.5cm]{figs/TwoSpheres.pdf} 
    \caption{(Left) The WCA potential is a minimum at r=2R. 
             (Right) The WCA potential increases rapidly for $r<2R$.}
    \label{fig:TwoSpheres}
\end{figure}

@@

Freezing is governed by entropy, which the universe %, or any closed system, 
seeks to maximize according to the second law of Thermodynamics. When the 
entropy of a pure substance in its liquid state is lower than the entropy 
that it would have in its solid state at the same temperature (necessary 
for thermal equilibrium), the same pressure (necessary for mechanical 
equilibrium), and the same chemical potential (necessary for diffusive 
equilibrium), an abrupt phase transition occurs, that is, the liquid 
freezes. The particular structure and density that result are affected 
by the interatomic potential. When repulsive, soft sphere fluids 
(that are not too soft) freeze, they form lattice structures composed of 
Face Centered Cubic (FCC) crystal cells~\cite{Hansen}, as do hard spheres. 
This structure is what one would see by filling a box with marbles. It 
results in the highest density achievable for hard, and nearly-hard, spheres 
that are purely repulsive.

\begin{figure}
    \centering
    \includegraphics[height=8cm]{figs/P-T_Diagram.pdf}
    \caption{Conceptual Pressure-Temperature Phase Diagram}
    \label{fig:P-T_Diagram}
\end{figure}

For pure substances, a phase transition occurs along a line on a 
Pressure-Temperature (P-T) phase diagram, such as can be seen in the phase 
diagram shown in Figure~\ref{fig:P-T_Diagram}. This phase diagram is 
representative of some arbitrary, unknown substance composed of atoms, 
or molecules which exhibit both attraction and repulsion. However, since 
the atoms in a WCA fluid do not exhibit attraction there will not be a 
dividing line between the gas and liquid phases, there will just simply
be a fluid, which we will refer to as a liquid in this paper.
%which may be referred to as a "liquid", since at the phase boundary the fluid is dense. 
Points along the line that separates the solid and the liquid phases 
indicate a phase transition where the solid and liquid phases coexist, 
and show that both the solid and the liquid have the same temperature 
and the same pressure at a phase transition.

The pressure can be related to the Helmholtz free energy F, which is minimized 
for a system at fixed temperature, volume and number of particles, as the 
entropy of the system and its surroundings is maximized. The negative of 
the partial derivative of the Helmholtz free energy with respect to volume V, 
and at a fixed temperature T and number of particles N, equals the pressure. 
\begin{equation}{P=-\frac{\partial{F}}{\partial{V}}\bigg|_{T,N}}\end{equation}
\noindent Dividing both the Helmholtz free energy and the volume by the 
number of particles, the partial derivative of the Helmholtz free energy 
with respect to volume becomes the partial derivative of the Helmholtz 
free energy per atom, $f$,  with respect to the inverse number density. 
\begin{equation}{P=-\frac{\partial{\frac{F}{N}}}{\partial{\frac{V}{N}}}\bigg|_{T,N} = -\frac{\partial{f}}{\partial{\frac{1}{n}}}\bigg|_{T,N}}\end{equation} 
Thus, the pressure corresponds to the negative of the slope of the curve 
on a plot of the Helmholtz free energy per atom versus the inverse 
density at fixed temperature.

 \begin{figure}
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig1.pdf}
    \caption{Liquid and solid densities at the solid-liquid phase 
    transition are identified using Maxwell's Double Tangent Construction}
    \label{fig:MaxwellDT}
  \end{figure}

Two such curves are shown in Figure~\ref{fig:MaxwellDT}, one for the liquid 
and one for the solid at a fixed temperature. Each curve consists of a 
collection of data points $(\frac{F}{N}, \frac{1}{n})$ for the liquid and
the solid respectively, which will be obtained by using cDFT to find the 
Helmholtz free energies of the liquid and the solid for various 
temperatures and number densities. 
Since the pressure of the 
liquid is the same as the pressure of the solid at the point of transition, 
the density of the liquid and the density of the solid at the point of 
freezing can be found from identifying the points at which the 
tangents to each curve have the same slope, as shown by the black line. 
This process is called Maxwell's Double Tangent Construction.

To facilitate finding the pressure at the phase transition computationally, 
it is helpful to consider the Gibbs free energy given by G=F+PV. 
Rearranging the terms and dividing by N gives an  expression for the 
Helmholtz free energy per atom $f$ in terms of the Gibbs free energy 
per atom  $g$ (which is the chemical potential)  
\begin{equation}\frac{F}{N} = -P\frac{V}{N}+\frac{G}{N} \end{equation}
\begin{align}f=-P\frac{1}{n}+g\end{align}
This equation has the form of a line
\begin{equation}
    y=mx+y_o
\end{equation}
and describes the black line shown in Figure~\ref{fig:MaxwellDT} with the 
pressure given by the negative of the slope of the line, and the Gibb's 
free energy per atom, or chemical potential given by the y-intercept. 
At the point of phase transition the chemical potential of the liquid is 
the same as the chemical potential of the solid, as is the pressure.
\begin{equation}
    \mu_L=\mu_S\mbox{~~~~and~~~~}P_L=P_S\mbox{~~~~at a phase transition}
\end{equation}

 \begin{figure}
    \centering
     \includegraphics[height=8cm]{figs/MaxwellDTC-Fig2.pdf}
    \caption{Pressure is found at point of intersection which occurs 
    when $\mu_L=\mu_S$.}
    \label{fig:GibbsvsP}
  \end{figure}

\begin{figure}
    \centering
    \includegraphics[height=8cm]{figs/MaxwellDTC-Fig3.pdf}
    \caption{The liquid and solid densities at the phase transition are 
    found from a plot of pressure vs inverse density for both the liquid and the solid.}
    \label{fig:Pvsinvn}
  \end{figure}

The pressure at the phase transition can be obtained from the point of 
intersection that occurs when the chemical potential 
is plotted against the pressure for both the liquid and the 
solid at fixed temperature, as shown in Figure~\ref{fig:GibbsvsP}. 
This plot is constructed from data collected from lines tangent 
along the curves on the Helmholtz free energy per atom verses inverse 
density plots (like that of Figure~\ref{fig:MaxwellDT}) for both the 
liquid and the solid at a given temperature. 
The slope of the tangent line at a given point gives the negative of the 
pressure at that point, and the corresponding y-intercept of the tangent 
line gives the chemical potential at that point. Collecting this data 
(P,~$\mu$) at each point for the liquid and the solid respectively, 
produces the curves in Figure~\ref{fig:GibbsvsP}.

\begin{figure}
    \centering
    \includegraphics[height=8cm]{figs/T-n_Diagram.pdf}
    \caption{Conceptual Temperature-Density Phase Diagram used to identify 
    the number densities of the liquid and the solid during a phase 
    transition. The gray region is where different phases coexist in 
    varying amounts at their phase transition number densities.}
    \label{fig:T-n_Diagram}
  \end{figure}

Once the pressure is known at the point of phase transition, for a given 
temperature and number density, the liquid and solid number densities at 
the phase transition can be found from plots of the pressure versus the 
inverse of the number density for both the solid and the liquid, 
as shown in Figure \ref{fig:Pvsinvn}. By repeating this procedure to 
find the liquid and solid number densities at the phase transition 
for various temperatures, a Temperature-Density (T-n) phase diagram 
can be generated, similar to that shown in Figure \ref{fig:T-n_Diagram}, 
but without the gray hump characteristic of fluids that have 
both attractive and repulsive interatomic potentials.
